---
title: "Text demo"
author: Casey Breen
---

Summary: In this notebook, we'll read in PDF files. 

```{r}
## Library packages (you'll have to install these if you haven't already)
library(tidyverse)
library(tesseract)
library(tidytext)
library(tidyverse)
library(pdftools)

## set engine for OCR
eng <- tesseract("eng")
```
## Images 

First, test this out with a .png or jpeg file. You can just give it the image URL directly. 

```{r}
## read in image 
text <- tesseract::ocr("http://jeroen.github.io/images/testocr.png", engine = eng)

## print out text
cat(text)
```
## PDF Files 

Now we'll work with a pdf file. First, we'll need to convert the PDF to an image file. 

```{r}
## convert to PDF
pngfile <- pdftools::pdf_convert('~/Downloads/service-ll-usrep-usrep095-usrep095670-usrep095670.pdf', dpi = 600)
```
Now, we can use the OCR function from tesseract. 

```{r}
## Perform OCR
text <- tesseract::ocr(pngfile)

## print out page 4
cat(text[4])
```

## Analysis

Finally, we can do some basic analysis of the text... 

```{r}
## First put it into a data.frame 
df <- tibble(doc = 1, text = text[1])

## split into one word (token) per row
tokens <- df %>% 
  tidytext::unnest_tokens(word, text)

## create count of most common words over 5 characters
words <- tokens %>% 
  count(word, sort = TRUE) %>% 
  filter(nchar(word) > 5)
```

Exercises: 

1. Find a new image from the internet and see how well we can process its text. 

2. Create a bar plot (geom_col) of the top 10 words over 5 letters in the 3rd page of the UNITED STATES v. Fox document. 

3. Bonus: Find a new PDF, scan it in, and find the top 10 words over 5 letters. How do they compare. 

